---
title: "Building Blocks Mapping"
author: "Anissa El Marrahi"
date: "3/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

After having done an Exploratory Data Analysis on tumors from the Metabric dataset (tumor samples and their proportions of 22 immune cell types obtained with the deconvolutiuon algorithm CIBERSORT), we found the structure of a simplex showing that tumors might have some archetypal behaviours regarding the immune system. It was visible with the 4 vertices of a 3D simplex, showing that some tumors might have special architectures/organization or a mix of those 4 typical organizations.

Now, we will try to map our observations with other results from Fabio Lipreri's (MSc in Computer Science,Karolinska Institutet & SciLifeLab) analyses. He worked on the cell organization of tumors from a batch of patients with Breast Cancer, data taken from the article [Keren et al., Cell (2018)](https://www.cell.com/cell/pdf/S0092-8674(18)31100-0.pdf) and obtained by multiplex imaging (MIBI TOF, mass cytometry Time Of Flight). He screened a set of sites with a given radius inside a plane showing the organization  of different cell types within a tumor. After a PCA of those sites on the cell abundance space, the result was a 3D simplex with 4 vertices. Each vertex has been identified by Archetypal Analysis to end up with building blocks of tumors. The closest points (sites) to one vertex have been classified as part of a building block. The ones which were in the middle of the figure were classified as a mix of those 4 building blocks.

Here, the aim is to try to see which immune cell types constitute each building blocks regarding gene markers in order to see if there's a correspondence with the results that we found after having observed the population of tumors.


## Data pre processing


First, let's charge the libraries needed for data processing and visualization
```{r libs}

rm(list=ls())
library(tidyverse)
library(ade4)
library(ggplot2)
library(reshape2)
library("factoextra")
library(plotly)
dirName <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(dirName)#setwd("/scratch/anissa.el/ImmuneStates")

```



```{r data, echo=TRUE,eval=FALSE}
#archetypes_sites <- read.csv('/srv/mfs/hausserlab/fabio/output/csv_files/archetypes_matrix.csv', header = TRUE, sep = ",",col.names = c('archetype1', 'archetype2', 'archetype3', 'archetype4','site_id','patient_id'))
#saveRDS(archetypes_sites,"/srv/mfs/hausserlab/anissa.el/ImmuneStates/data_BB/archetypesSitesKeren.rds")
#sites_abund <- read.csv("/srv/mfs/hausserlab/fabio/output/csv_files/abbundance_matrix.csv",header = TRUE, sep = ",",col.names = c("index",'CD8-T', 'Other immune', 'DC / Mono', 'CD3-T', 'B', 'NK', 'Keratin-positive tumor', 'Tumor', 
#              'CD4-T', 'Mesenchymal-like', 'Macrophages', 'Endothelial', 'Tregs', 'Unidentified', 'DC', 'Mono / Neu', 
#              'Neutrophils','site_id', 'patient_id'))
#sites_abund <- dplyr::select(sites_abund,-(index))

```


```{r dataGaussian, eval=TRUE,echo=TRUE}

NBCELLTYPES = 17
CELLTYPES <- c('CD8-T', 'Other immune', 'DC / Mono', 'CD3-T', 'B', 'NK', 'Keratin-positive tumor', 'Tumor','CD4-T', 'Mesenchymal-like', 'Macrophages', 'Endothelial', 'Tregs', 'Unidentified', 'DC', 'Mono / Neu','Neutrophils')
archetypes_sites <- read.csv("./data_BB/archetypes_matrix_gaussian.csv", header = TRUE, sep = ",")#,col.names = c('archetype1', 'archetype2', 'archetype3', 'archetype4','site_id','patient_id'))

saveRDS(archetypes_sites,"./data_BB/archetypesSitesKeren.rds")
#saveRDS(archetypes_sites,"/srv/mfs/hausserlab/anissa.el/ImmuneStates/archetypesSitesKeren.rds")
sites_abund <- read.csv("./data_BB/abundance_matrix_gaussian.csv",header = TRUE, sep = ",")#,col.names = c("index",'CD8-T', 'Other immune', 'DC / Mono', 'CD3-T', 'B', 'NK', 'Keratin-positive tumor', 'Tumor','CD4-T', 'Mesenchymal-like', 'Macrophages', 'Endothelial', 'Tregs', 'Unidentified', 'DC', 'Mono / Neu','Neutrophils','site_id', 'patient_id'))
sites_abund <- sites_abund%>%dplyr::select(-(index))

```

  We have a dataframe with all the coordinates of the 3D PCA made in  the previous analysis in the cell abundance space + 3 columns (arch1_prob to arch4_prob) representing the probability to belong to one of the 4 archetypes. This will help us to allocate the sites into our 4 building blocks.


### Allocate to each building block a set of sites

  Here, we decided to select the 10 % closest sites of each building block. But the selection will based on the alpha probabilities taken from the archetypal analysis. Indeed, for each site, a vector alpha of size 4 have been computed. It represents the probabilities to respectively belong to archetype 1, 2 3, 4. SO the sum of the elements of this vector is equal to 1 because we don't consider the probability to not belong to at least one of the archetypes as they fit to our dataset. So instead of using the space to select the sites that correspond the most to the building blocks (we don't take a look at the sites with mixed archetypes), we select the sites for which the probability to belong to one of the archetypes is superior to 0.9.

```{r 10sites, echo=TRUE}

block1 <- arrange(archetypes_sites, desc(archetype1))%>%filter(archetype1>=0.9)
block2 <- arrange(archetypes_sites, desc(archetype2))%>%filter(archetype2>=0.8)
block3 <- arrange(archetypes_sites, desc(archetype3))%>%filter(archetype3>=0.9)
block4 <- arrange(archetypes_sites, desc(archetype4))%>%filter(archetype4>=0.9)
#block4
```

### Link the building blocks with cell types


```{r joints, echo=TRUE}

#sites <- relocate(sites,site_ID,patient_ID)
#bblock1_sites <- filter(sites_abund, site_id %in% block1$site_id & patient_id %in% block1$patient_id)
bblock1_sites <- semi_join(sites_abund,block1,by = c('site_id','patient_id'))

bblock2_sites <- semi_join(sites_abund,block2,by = c('site_id','patient_id'))

bblock3_sites <- semi_join(sites_abund,block3,by = c('site_id','patient_id'))

bblock4_sites <- semi_join(sites_abund,block4,by = c('site_id','patient_id'))


```

### Visualization of the sites (cell abundance)

Let's take a look at the cell abundance for each building block 


#### Building Block 1

```{r plot1, echo=TRUE}

bb1 <- bblock1_sites[1:NBCELLTYPES]
#corrBB1 <- cor(bb1,method="pearson")
#heatmap(x = corrBB1, symm = TRUE)
summary(bb1)
colnames(bb1)
bb1_1 <- bb1 %>% mutate(Other = Unidentified + Mesenchymal.like,.keep = "unused") %>%
  mutate(Epithelial = Tumor + Keratin.positive.tumor,.keep = "unused")%>%
  mutate(Monocytes = Mono...Neu+DC...Mono,.keep = "unused")%>%
  mutate(OtherImmune = Other.immune+Neutrophils +CD3.T,.keep = "unused")

if(rowSums(bb1)!=rep.int(1,nrow(bb1))){
  bb1_1 <- bb1_1/rowSums(bb1_1)
  bb1 <- bb1/rowSums(bb1)
}

bb1_m <- melt(bb1)
ggplot(bb1_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 45,hjust = 0.8,vjust = 0.5))+
  labs(title = "    Boxplot of proportions of cell types from Building Block 1 sites")+
  xlab ("") + ylab("proportion")


```

The majority of the sites landscapes are composed of B cells and a little amount of CD4 T cells



#### Building Block 2

```{r plot2, echo=FALSE}

bb2 <- bblock2_sites[1:NBCELLTYPES]
bb2_1 <- bb2 %>% mutate(Other = Unidentified + Mesenchymal.like,.keep = "unused") %>%
  mutate(Epithelial = Tumor + Keratin.positive.tumor,.keep = "unused")%>%
  mutate(Monocytes = Mono...Neu+DC...Mono,.keep = "unused")%>%
  mutate(OtherImmune = Other.immune+Neutrophils +CD3.T,.keep = "unused")

if(rowSums(bb2)!=rep.int(1,nrow(bb2))){
  bb2_1 <- bb2_1/rowSums(bb2_1)
  bb2 <- bb2/rowSums(bb2)
}

bb2_m <- melt(bb2)
ggplot(bb2_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 45,hjust = 0.8,vjust = 0.5))+
  labs(title = "    Boxplot of proportions of cell types from Building Block 2 sites")+
  xlab ("") + ylab("proportion")

```

  We can see here a big heterogeneity of cell types. The main part of them are CD4 T cells and CD8 tumors, but also Natural Killers cells. We can find also some macrophages (<20 %) and of course tumor cells (Keratin-positive tumor cells). It is quite logical to find certain immune cells such as CD8 Tm NK and macrophages as they globally have the same type of mechanism : they are tumor suppressors. FOr macrophages, it's more complicated as this cell type can adopt polarized behaviors regarding the signals and cytokines ==> tumor suppressor or tumor enhancer with M2 macrophages. This will be discussed in further analyses.

#### Building Block 3

```{r plot3, echo=FALSE}

bb3 <- bblock3_sites[1:NBCELLTYPES]
if(rowSums(bb3)!=rep.int(1,nrow(bb3))){
  bb3 <- bb3/rowSums(bb3)
}

bb3_m <- melt(bb3)
ggplot(bb3_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 45,hjust = 0.8,vjust = 0.5))+
  labs(title = "    Boxplot of proportions of cell types from Building Block 3 sites")+
  xlab ("") + ylab("proportion")

```

  In this building block, we mainly find tumor cells. If most of the sites here don't have significant amounts of cell types , we can make the hypothesis that we have cold tumors (tumors with low amount of TILs = Tumor Infiltrating Lymphocytes).


#### Building Block 4


```{r plot4, echo=FALSE}

bb4 <- bblock4_sites[1:NBCELLTYPES]

if(rowSums(bb4)!=rep.int(1,nrow(bb4))){
  bb4 <- bb4/rowSums(bb4)
}

bb4_m <- melt(bb4)
ggplot(bb4_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 45,hjust = 0.8,vjust = 0.5))+
  labs(title = "    Boxplot of proportions of cell types from Building Block 4 sites")+
  xlab ("") + ylab("proportion")
```

  This Building block is a mix of many cell types but we have dense boxplots also because of the density of this building block (238 sites, while the other building blocks are composed of less than 30 sites). We have a lot of mesenchymal-like cells and again tumor cells. But we can also find proportions of cells identified as "Other Immune cells" ==> we have to look at the classification methods used to see if we can identify some well-known cells. We could find again some macrophages.


```{r getData, echo=FALSE}

bb1 <- bb1 %>%mutate(bb = "bb1")
bb2 <- bb2 %>%mutate(bb ="bb2")
bb3 <- bb3 %>%mutate(bb = "bb3")
bb4 <- bb4 %>%mutate(bb = "bb4")
BuildingBlocks <- rbind(bb1,bb2)
BuildingBlocks <- rbind(BuildingBlocks,bb3)
BuildingBlocks <- rbind(BuildingBlocks,bb4)
BuildingBlocks <- BuildingBlocks %>%mutate(`Mesenchymal.like` =`Mesenchymal.like`+Unidentified)%>%dplyr::select(-(Unidentified))
#BuildingBlocks <- BuildingBlocks%>%column_to_rownames(var="bb")
write.csv(BuildingBlocks,"./data_BB/building_blocks_cell_abundance.csv",col.names=TRUE)
```


### Cell types markers

  For  each site  of one building block, we'll take a look at each cell, and the intensity value of 40 markers (more or less) detected by mass cytometry combined with imaging techniques in order to localize those cell types in a 2D space (tumor). Here, the aim is to compare and look for markers that can allow us to give a more specific description of the cell type. This, in order to find similarities with the analysis on tumors population (immune cell types determined by CIBERSORT algorithm on bulk RNAseq data from tumors of the METABRIC dataset).


```{r cellmarkers, echo=FALSE}

cell_markers <- read.csv('/srv/mfs/hausserlab/fabio/david_data/cellData.csv',header = TRUE, sep =',')

colnames(cell_markers)
cell_markers <- cell_markers %>% relocate(Background, .after = cellLabelInImage)
cell_markers <- dplyr::select(cell_markers,-(cellSize:Fe))
#SampleID,cellLabelInImage,cell_markers,Pan.Keratin,Keratin6,Keratin17,p53,Beta.catenin,EGFR,
#                        dsDNA,H3K9ac,H3K27me3,phospho.S6,Ki67,
#                        CD45,CD3,CD8,CD4,FoxP3,CD20,CD45RO,CD56,CD16,CD138,CD11b,CD11c,CD68,CD63,MPO,HLA.DR,HLA_Class_1,CD209,
#                        Lag3,PD1,PD.L1,IDO,
#                        Vimentin,SMA,CD31,
#                        B7H3,OX40,CD163,CSF.1R,tumorYN,tumorCluster,Group,immuneGroup

cell_markers <- dplyr::select(cell_markers,-c( immuneCluster, Ta, Au))%>% 
  mutate(immuneGroup = recode(immuneGroup,`0`= 'None',`1`='Tregs', `2`='CD4-T',
                              `3`='CD8-T', `4`='CD3-T', `5`='NK',
                    `6`='B', `7`='Neutrophils', `8`='Macrophages', `9`='DC',
                    `10`='DC / Mono', `11`='Mono / Neu', `12`='Other immune')) %>% 
  mutate(Group = recode(Group,`1`='Unidentified', `2`='Immune',
                        `3`='Endothelial', `4`='Mesenchymal-like',
                        `5` = 'Tumor',
                        `6` = 'Keratin-positive tumor'))

cell_markers <- cell_markers%>%
  mutate(cell_type = ifelse(Group == 'Immune', cell_type<- immuneGroup,cell_type <- Group))%>%
  dplyr::select(-(tumorYN:immuneGroup))

cell_markers <- cell_markers%>%
  dplyr::rename(patient_id = SampleID)%>%
  dplyr::rename(cell_id = cellLabelInImage)
colnames(cell_markers)
#ImmuneCells <- filter(cell_markers, immuneGroup !='None')
saveRDS(cell_markers, "./data_BB/cell_markers_keren.rds")
cells_abund_patient <- cell_markers %>%
  dplyr::select(patient_id,cell_id,cell_type)
cells_abund_patient <- cells_abund_patient%>%
  group_by(patient_id,cell_type)%>%
  tally()%>%
  pivot_wider(names_from = cell_type,values_from=n)
cells_abund_patient <- cells_abund_patient%>%
  replace(is.na(.),0)%>%
  column_to_rownames(var="patient_id")

CellsPropKeren <- t(apply(as.matrix(cells_abund_patient),1,function(x) x*100/sum(x)))

saveRDS(CellsPropKeren, file = "./wagner_analysis/data/CellsPropKeren.rds")

```



```{r pcaCellGrain, echo=FALSE}
cell_markers3 <- cell_markers %>%group_by(patient_id,cell_id)%>%mutate(nb = n())%>%ungroup()%>%dplyr::select(patient_id, cell_id, nb,cell_type)%>%pivot_wider(id_cols = c(patient_id,cell_id), names_from = cell_type,values_from = nb)%>%mutate_all(.,~ifelse(is.na(.),0,.))

sitesGrained.pca <- dudi.pca(cell_markers3%>%dplyr::select(-c(patient_id,cell_id)), center=TRUE,scale=FALSE,scannf=FALSE,nf=3)
fviz_eig(sitesGrained.pca)
fviz_pca_biplot(sitesGrained.pca, label="var",invisible = "ind")

fig1 <- plot_ly(x = sitesGrained.pca$li$Axis1, 
                y = sitesGrained.pca$li$Axis2,
                z = sitesGrained.pca$li$Axis3,
                #color =~cell_markers3$,
                showlegend = TRUE,
                type = "scatter3d", mode = "markers",
                marker = list(symbol = "triangle", size = 6, color = toRGB("steelblue", .7)),
                mode = "text",
                name = "tumor")
fig1 <- fig1 %>% layout(scene = list(xaxis = list(title = "PC1"), 
                                     yaxis = list(title = "PC2"), 
                                     zaxis = list(title = "PC3") ),
                        title = "PCA on cells sites space from TNBC")
fig1



markers1 <- cell_markers[,c("dsDNA","Background","Vimentin","SMA","B7H3","FoxP3","Lag3","CD4","CD16","CD56","OX40","PD1","CD31","PD.L1","EGFR","Ki67","CD209","CD11c","CD138","CD163","CD68","CSF.1R","CD8","CD3","IDO","Keratin17","CD63","CD45RO","CD20","p53","Beta.catenin","HLA.DR","CD11b","CD45","H3K9ac","Pan.Keratin","H3K27me3","phospho.S6","MPO","Keratin6","HLA_Class_1")]
markersPCA <- dudi.pca(markers1, center=FALSE, scale=FALSE, scannf=FALSE, nf = 3 )
fviz_eig(markersPCA)
fviz_pca_biplot(markersPCA, col.var = 'firebrick3',invisible = "ind",label= "var",alpha.ind=0.3,repel=TRUE)
fig2 <- plot_ly(x = markersPCA$li$Axis1, 
                y = markersPCA$li$Axis2,
                z = markersPCA$li$Axis3,
                color =~cell_markers$Vimentin,
                showlegend = TRUE,
                type = "scatter3d", mode = "markers",
                marker = list(symbol = "triangle", size = 6),#, color = toRGB("steelblue", .7)),
                mode = "text",
                name = "tumor")
fig2 <- fig2 %>% layout(scene = list(xaxis = list(title = "PC1"), 
                                     yaxis = list(title = "PC2"), 
                                     zaxis = list(title = "PC3") ),
                        title = "PCA on cells sites space from TNBC")
fig2

#check_cells <- right_join(CellsSitesIC%>%group_by(patient_id,site_id,cell_id)%>%summarise(counts = n()),cell_markers, by=c("patient_id","cell_id"))

```

Selection of markers in each cell according to sd(markers) cell-type-wise. Criteria, sd>2 ==> we consider those markers of interest as highly variable so likely to be explained by the environment (or not).

```{r selectMarkers, echo=TRUE}
markersPerCell.sd <- cell_markers%>%pivot_longer(cols = c("dsDNA","Background","Vimentin","SMA","B7H3","FoxP3","Lag3","CD4","CD16","CD56","OX40","PD1","CD31","PD.L1","EGFR","Ki67","CD209","CD11c","CD138","CD163","CD68","CSF.1R","CD8","CD3","IDO","Keratin17","CD63","CD45RO","CD20","p53","Beta.catenin","HLA.DR","CD11b","CD45","H3K9ac","Pan.Keratin","H3K27me3","phospho.S6","MPO","Keratin6","HLA_Class_1"),names_to = "marker", values_to= "value")%>% group_by(cell_type, marker)%>%summarise(sd = sd(value))


ggplot(data = markersPerCell.sd ,aes(x=cell_type,y = sd,group = marker))+
  geom_point()+
  geom_hline(yintercept=2, linetype="dashed", color = "red")+
  theme(axis.text.x = element_text(angle = 90, hjust = 0.2, vjust = 0.2))+
  labs(x = "")

selectedMarkers <- markersPerCell.sd%>%filter(sd >=2)#%>%ungroup()%>%dplyr::select(marker)%>%unique()

```


Here, we get the cell ID and site ID to link each cell to a site and so on :

```{r cellsmarkers2, echo= TRUE}

cells_per_sites<-read.csv("./data_BB/site_cells_gaussian.csv",header = TRUE, sep = ",") #read.csv("/srv/mfs/hausserlab/fabio/output/csv_files/site_cells.csv",header = TRUE, sep = ",")
cells_per_sites <- cells_per_sites %>% dplyr::rename (site_id = site_idx)

```


We create our 4 blocks of cells and join them with the table of cell markers to have all the cells from one building block and their values for every marker: 

```{r cellsblocks, echo = FALSE}

CellsBB1 <- semi_join(cells_per_sites,bblock1_sites,by = c('site_id','patient_id'))
CellsBB1_markers <- semi_join (cell_markers,CellsBB1, by = c('cell_id','patient_id'))


CellsBB2 <- semi_join(cells_per_sites,bblock2_sites,by = c('site_id','patient_id'))
CellsBB2_markers <- semi_join (cell_markers,CellsBB2, by = c('cell_id','patient_id'))

CellsBB3 <- semi_join(cells_per_sites,bblock3_sites,by = c('site_id','patient_id'))
CellsBB3_markers <- semi_join (cell_markers,CellsBB3, by = c('cell_id','patient_id'))

CellsBB4 <- semi_join(cells_per_sites,bblock4_sites,by = c('site_id','patient_id'))
CellsBB4_markers <- semi_join (cell_markers,CellsBB4, by = c('cell_id','patient_id'))


```


### **Markers of B cells and CD4 cells in Building Block 1** 

  Let's take a look at only B cells and CD4 T cells as they seem to represent this building block (in higher proportions). In other words, we are going to see what are the main markers of those cell types within this building block and try to map those cell types with our findings.
  
```{r functplot, echo = FALSE}
# This function creates a boxplot of a set of markers (cell, cancer and stroma markers have been excluded) for a chosen celll type from a building block and prints It
markers_BB_boxplot <- function (BBMarkers,nameBB, cellType,scaleIntensity){
  cell_BB <- filter(BBMarkers,cell_type == cellType)%>%
    dplyr::select(patient_id,cell_id,cell_type,Background,CD45,CD3,CD8,CD4,FoxP3,CD20,CD45RO,CD56,CD16,CD138,CD11b,CD11c,CD68,CD63,MPO,HLA.DR,HLA_Class_1,CD209,Lag3,PD1,PD.L1,IDO,B7H3,OX40,CD163,CSF.1R)

cell_BB <- cell_BB%>%dplyr::select(-c(CD11b,CD45))
cell_BB_m <- melt(cell_BB[3:27])

ggplot(cell_BB_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE,outlier.colour=NA) + 
  coord_cartesian(ylim = scaleIntensity)+
  theme(axis.text.x=element_text(angle=45,hjust=0.8,vjust=0.5))+
  labs(title = paste("    Boxplot of cell markers intensities for ",cellType," from ",nameBB))+
  xlab ("") + ylab("intensity (Arbitrary Unit)")
}

#THis function is the same as the first one, the only difference is that we included markers of tumor cells + marker of cell growth : Ki67
markers_BB_boxplot2 <- function (BBMarkers,nameBB, cellType,scaleIntensity){
  cell_BB <- filter(BBMarkers,cell_type == cellType)%>%
    dplyr::select(patient_id,cell_id,cell_type,Background,CD45,CD3,CD8,CD4,FoxP3,CD20,CD45RO,CD56,CD16,CD138,CD11b,CD11c,CD68,CD63,MPO,HLA.DR,HLA_Class_1,CD209,Lag3,PD1,PD.L1,IDO,B7H3,OX40,CD163,CSF.1R,Pan.Keratin,Keratin6,Keratin17,p53,Beta.catenin,EGFR,Ki67,Vimentin,SMA)

cell_BB <- cell_BB%>%
  dplyr::select(-c(CD11b,CD45))
cell_BB_m <- melt(cell_BB[3:37])
ggplot(cell_BB_m,aes(color = variable)) +
  geom_boxplot(aes(x = variable,y = value),show.legend = FALSE,outlier.colour=NA) + 
  coord_cartesian(ylim = scaleIntensity)+
  theme(axis.text.x=element_text(angle=45,hjust=0.8,vjust=0.5))+
  labs(title = paste("    Boxplot of cell markers intensities for ",cellType," from ",nameBB))+
  xlab ("") + ylab("intensity (Arbitrary Unit)")
}

```



#### B cells

```{r plotBCells, echo=FALSE}

markers_BB_boxplot(CellsBB1_markers,"BB1","B",c(0,7))
ggsave("../figs/markersBcellsBB1.pdf",height = 3, width = 4)
```
 
 
 MIBI stands for Multiplex Ion Beam Imaging

A principal Component Analysis might be helpful to get information about the markers of some sub-types of B cells :

```{r pcaBcells, echo = FALSE}


Bcells_bb1 <- CellsBB1_markers %>% filter(cell_type == 'B') %>%
  dplyr::select(patient_id,cell_id,cell_type,CD20,CD56,CD16,CD138,CD68,
                CD63,MPO,HLA_Class_1,CD209,Lag3,IDO,B7H3,OX40,CD163,
                CSF.1R,PD1,PD.L1,CD4,FoxP3,CD11c,CD45RO,CD3,CD8)
#-(CD11b,,HLA.DR,CD45,)
BB1_Bcells_PCA <- dudi.pca(Bcells_bb1[4:24],scale = FALSE,scannf = FALSE,nf =3)

fviz_eig(BB1_Bcells_PCA)

fviz_pca_biplot(BB1_Bcells_PCA,col.var = 'indianred2',col.ind = 'steelblue2',invisible = 'ind', repel = TRUE)

fviz_pca_var(BB1_Bcells_PCA,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

```
 
* CD45RO marker is a cell surface marker of T helper cells.
* CD45 is a widely known marker used to identify immune cells.
* CD11b is a marker of neutrophils.
* HLA-DR is a marker for monocytes.
 
 
 We removed those markers because we only took B cells. And we know that B cells belong to the immune cells group, monocytes group and so on. We already removed the markers of cells and those from tumor cells to shrink our range of markers. Indeed, It can be complicated to have a lot of parameters that are not every time relevant and therefore can act like noise or hide some effects of certain markers that are in reality very important.
 
 **WARNING** : There are markers that are well-known to identify better some cell types such as **Ly6C** which is commonly used in flow cytometry to label monocytes. To identify deeply some cell types, there must be a combination of markers that are used.
 
 
 In this PCA, we set the parameter scale on FALSE because we don't want to take in account the "noise" that other markers from different immune cell type might produce in our analysis. For example, some markers more related to T cells might have higher intensities but they are not related to B cells so they can be considered as noise. Hence, we don't scale our data.
 
 What we find  on the markers that contribute the most to our dataset is quite relevant. Indeed, we already know that CD20, CD63 are related to B cells. Let's see If we can find some sub-types such as plasma cells or B cell memory sub-type :
 
```{r pcaBcells2, echo = FALSE}

# 
# CD138_intensity <- Bcells_bb1$CD138
# fig <- plot_ly(x = BB1_Bcells_PCA$li$Axis1, y = BB1_Bcells_PCA$li$Axis2, z = BB1_Bcells_PCA$li$Axis3, type="scatter3d", mode="markers",
#                color = ~CD138_intensity,mode = 'text')
# 
# fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'), yaxis = list(title = 'PC2'), zaxis = list(title = 'PC3') ),
#                       title = '3D PCA of cell markers on B cells from building block 1 (with CD138 intensity)')
# fig
 
```
 
We don't find this marker in higher intensities in the group of B cells. This is a good marker for plasma cells but apparently, there doesn't seem to be any plasma cells inside the group, or we need to test other more specific markers to determine the presence/absence of plasma cells.
For memory cells, according to [R&D website](https://www.rndsystems.com/resources/cell-markers/immune-cells/b-cells/memory-b-cell-markers), CD27 marker combined with the presence of CD19 marker are used to identify B cell memory. 

So here, we can't determine accurately the sub-types of B cells.


#### CD4 cells in building block 1

Let's see with a boxplot the markers that are most intense :

```{r plotCD4BB1, echo=FALSE}

markers_BB_boxplot(CellsBB1_markers,"BB1","CD4-T",c(-2,7))
ggsave("../figs/markersCD4TBB1.pdf",height = 3, width = 4)
```
 We can see that in majority CD4, CD3 and CD45RO positive markers. We didn't find a significant intensity for FoxP3, nor checkpoint markers such as B7H3 (CD276), PD1 and Lag3. 
 
The boxplot of CD45RO markers seems to be centered with a lot of CD4 cells whom intensity is near the median. Maybe, It is a sign that we have some T helper cells.


#### What about tumor cells ?

The fact that immune system interacts with tumor micro environment has already been proven although It is still unclear. Here, we are going to take a look at the relationship between markers from tumor cells and the markers expressed only in immune cells. We added also Ki-67 that is very common marker to describe cell growth :

```{r boxplotTUmMarkers, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Kreatin positive tumor cell
#'Kreatin-positive tumor' %in%CellsBB1_markers$cell_type
markers_BB_boxplot2(CellsBB1_markers,"BB1",'Keratin-positive tumor',c(-1,5))
```
Here, we only have 2 cells in this site that are keratin positive tumors so our analysis is not really accurate.

Keratin 6 and 17 are markers of a worse prognosis in Breast Cancer. Those proteins that are part of Intermediate Filaments constitute the cytoskeleton. The are also involved in cell protection, particularly in cases of wounds where in healthy tissue, are up regulated (Keratin 6 and 17). We can also see the presence of CD45RO in low intensities, also for MHC class II marker.


### **Building block 2 : Macrophages, CD8 and CD4 cells**

If we take a look at the intensities for macrophages from this building block, we can find that some markers are highly intense :

```{r plotMacro, echo=FALSE}

markers_BB_boxplot(CellsBB2_markers,"BB2","Macrophages",c(-0.5,9))
ggsave("../figs/markersMacrophagesBB2.pdf",height = 3, width = 4)
```
 
 In particular, the marker CD68 seems significantly intense while we have low intensity (even no intensity) for the marker CD63, known to be a marker for M2 macrophages in flow cytometry. Therefore, we can identify the macrophages as more likely to be M1-like macrophages. W need to prove It with a statistical test. BUt, as we might have seen in this building block, the sites contain a lot of CD8+ T cells, CD4 T cells which are known to be tumor suppressors ==> Th1 response against cancer cells (+ Little amounts of NK cells). The CSFR1 marker is present because it is the main marker of monocytes.
 
 

There are few cells with a high amount of the CD63 marker while we know that the intensity of this marker can go up to 12/15.Here, the maximum is near 4. 

However, some articles say that CD63 is a common marker for a lot of cells that do exocytosis (exosomes) so maybe this marker is not really relevant here.
It is important to say that CD68 is used to be a marker of M1 macrophages but we don't know about M0 macrophages.

#### CD8 T cells  sub-classification 

```{r plotCD8, echo=FALSE}

markers_BB_boxplot(CellsBB2_markers,"BB2","CD8-T",c(-1,7))

```
 
 Unsurprisingly, the CD8 marker is highly intense, and CD3 too. We can say also that the CD45RO marker is expressed but in lower intensity (still higher than the background). The presence of HLA class I and II shows a relationship between CD8+ T cells which are made to kill foreign components of the organism and APC (Antigen Presenting Cells) and in general all nucleated cells (HLA class I). We can interpret that also like an approximation  of the MIBI-TOF analysis pipeline : As CD8+ bind to HLA receptors to release perforins/granzymes in malignant cells, sometimes the segmentation of the cells might be wrong. 



#### CD4+ T cells  sub-classification

 Let's see if we can refine the CD4-T class the same way :
 
```{r plotCD4, echo=FALSE}

markers_BB_boxplot(CellsBB2_markers,"BB2","CD4-T",c(-1,8))
ggsave("../figs/markersCD4TBB2.pdf",height = 3, width = 4)
```

 We can find the CD3 and CD4 markers in higher intensities but also a certain amount of CD8 marker. THis means that we are most likely to have a T helper cell but we cannot determine the subtype of those CD4 cells, we need more specific markers.


#### What about tumor cells ?

In this building block, we have 305 Keratin positive tumor cells

```{r boxplotTUmMarkers2, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Kreatin positive tumor cell
sum(bblock2_sites$Keratin.positive.tumor)
'Keratin-positive tumor' %in%CellsBB2_markers$cell_type
'Tumor' %in%CellsBB2_markers$cell_type
markers_BB_boxplot2(CellsBB2_markers,"BB2",'Keratin-positive tumor',c(-1,8.5))
```
 
 We can see a different profile than the same cells from building block 1. Here, we can see that the marker p53 is moderately  intense. It is a marker of mutations in tumor cells. There is also the marker Ki 67 but its median intensity is near 0 which is coherent with the immune profile in this building block that is more Th1 response-oriented. Here, the immune cells down-stimulate cell growth by killing the cells. It could be interesting to take a look at the survival for the patients from who we took the samples and see if the prognosis is better also. Even though we are looking at the keratin positive tumor cells here, the intensity of keratin markers is not as high as what we observed in the first building block. As we know that this building blocks seems to be more likely to represent a Th1 response, we can suppose that Keratin proteins might act as a defense of tumor cells against immune cells. They could act as a protection from the injuries of immune system. Or they can try to escape from the immune system by down regulating It with Th2 responses. Those hypotheses that have to be verified. 
 
 
### Building Block 4
 
```{r plotMacro5, echo=FALSE}

markers_BB_boxplot(CellsBB4_markers,"BB4","Macrophages",c(-1,3))

```

We will try to see if we can identify some M2 macrophages that are more likely to be CD209+ and CD163+


 We can also check the 'Other Immune cells' class :
 
```{r plotImmCell, echo=FALSE}

markers_BB_boxplot(CellsBB4_markers,"BB4","Other immune",c(-1,2.5))
```
The markers used can't allow us to sub-classify the 'Other Immune' type as the values of intensity are even lower than background intensities.
We can't conclude on the specific subtypes inside building blocks. Our dataset lacks specificity on the markers due to technical limitations. 


#### Tumor cells in Building block 4

```{r boxplotTumMarkers3, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Keratin positive tumor cell
sum(bblock4_sites$Keratin.positive.tumor)
sum(bblock4_sites$Tumor)
'Keratin-positive tumor' %in%CellsBB4_markers$cell_type
'Tumor' %in%CellsBB4_markers$cell_type
markers_BB_boxplot2(CellsBB4_markers,"BB4",'Keratin-positive tumor',c(-2,6))
```

Here, we have in total 1886 Keratin positive tumor cells.

We can see that we have big intensities of beta catenin and of course some keratin markers.
**Pan Keratin** is a marker  which stands for Cytokeratin AE1/AE3. This marker detects a group of keratins : 1, 8, 10, 14, 16 and 19 (From [Pathology outlines website](https://www.pathologyoutlines.com/topic/stainsae1ae3.html) )

Let's see the markers for cells simply labelled as "Tumor" (we have 321 of them in this building block): 

```{r boxplotTumMarkers4, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Keratin positive tumor cell
sum(bblock4_sites$Tumor)
'Tumor' %in%CellsBB4_markers$cell_type
markers_BB_boxplot2(CellsBB4_markers,"BB4",'Tumor',c(-2,2))
```

The most intense marker here is Beta-catenin which ,unsurprisingly, is a marker of tumors. We can find also HLA class I in very low (negative) intensities


```{r boxplotMesenchymalBB4, echo = FALSE}
mesenchymalCellsBB4 <- filter(CellsBB4_markers,cell_type == 'Mesenchymal-like')%>%arrange(desc(Ki67))%>%dplyr::select(Ki67)
markers_BB_boxplot2(CellsBB4_markers,'BB4', 'Mesenchymal-like', c(-1,3))
```

We have higher intensities for Vimentin and SMA, two main markers of mesenchymal like cells, they are markers of differentiation of some epithelial cells.
  
  
### Building Block 3 : is this building block really pure from any immune cell type ?

As we have already seen It, the building block 3 seems to be more tumor cells-oriented. However, we know that the immune system can sometimes be "hacked" by tumor cells to allow them grow. This is why we could observe in TIL, some T cells and macrophages that have polarized actions : pro-anti- tumor. We already know for example that M2 macrophages are involved in wound healing processes which explain the presence of endothelial cells or fibroblasts because of angiogenesis in favor of the tumor initiated by those immune cells paradoxically. 
We can't find any cell with a high intensity of the CD163 marker, so our hypothesis to find M2 macrophages in disorganized tissue is not true, or at least cannot be verified.



#### Do the tumor cells have a different profile ?

Let's take a look at tumor cells and their related markers

```{r boxplotTumMarkersBB3, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Kreatin positive tumor cell
#'Kreatin-positive tumor' %in%CellsBB3_markers$cell_type
'Tumor' %in%CellsBB3_markers$cell_type
markers_BB_boxplot2(CellsBB3_markers,"BB3",'Tumor',c(-2,4))
```

This seems to be a regular kind of tumor cell with a high intensity of beta catenin marker which is representative of a tumor cell behavior.

Let's take a look at Keratin positive tumor cells in this building block :

```{r boxplotTumMarkersBB3TUmors, echo = FALSE}
## NO TUMOR CELL TYPE HERE : Only Kreatin positive tumor cell
'Keratin-positive tumor' %in%CellsBB3_markers$cell_type
#'Tumor' %in%CellsBB3_markers$cell_type
markers_BB_boxplot2(CellsBB3_markers,"BB3",'Keratin-positive tumor',c(-2,7.5))
ggsave("../figs/markersKerPosCellsBB3.pdf",height = 3, width = 4)
```

THose cells express more pan keratin and keratin 17 and almost no keratin 6.

```{r boxplotKI67, echo = FALSE}
mesenchymalCellsBB3 <- filter(CellsBB3_markers,cell_type == 'Mesenchymal-like')%>%arrange(desc(Ki67))%>%dplyr::select(Ki67)
markers_BB_boxplot2(CellsBB3_markers,'BB3', 'Mesenchymal-like', c(-2,2.5))

```
The beta catenin seems to be very high in this cell type  but htis is the ionly marker htat seems to be expressed here.


### **What about mesenchymal like cells ?**

We know that there are cells at an undifferentiated stage in the tumor micro environment. It would be interesting to take a look at the markers related to those cells. But first, we need to see in which building block(s), this cell type is mostly present (in terms of proportions)

```{r mesenchymalCells, echo = FALSE }
# summary(bblock1_sites)
# summary(bblock2_sites)
# summary(bblock3_sites)
# summary(bblock4_sites)
```

We'll take a look at mesenchymal cells in building block 4 (median = 48)

```{r mesenchymalCllsBB4, echo = FALSE}
markers_BB_boxplot2(CellsBB4_markers,'BB4','Mesenchymal-like',c(-1,4))
```
 This cell type has negative intensities for beta catenin and HLA class I. In some websites, this last marker is presented a one for identifying this cell type but we don't understand why we find low intensities of this marker here. source('data_proc_building_blocks.r')
 
### **Are the results consistent ?**

We will run 3 times the simulation  of building blocks analysis to see if the cell abundance results that we have -especially for building block 4 which represents a disorganized tissue- is consistent  because the boxplots represented a lot of outliers  



### **Immune Checkpoint markers**

#### **Are they represented by a mix of building blocks ?**

Immune Checkpoints are critical for tumor progression but are also some interesting targets in immunotherapy. Indeed, they are response for tumor survival and correlate with low survival for patients. The major markers of immune checkpoints are Lag3, PD1 (T cells), PD-L1 (tumor cells) and B7H3. We didn't find any of those markers in our building blocks which can either indicate that they are not expressed in our set of cells or just that the cells expressing those markers are not representative of any of the 4 building blocks.  Then we can make the hypothesis that the cells expressing those IC are part of a mix of building block. 

To test this hypothesis, we would take a look first at the cell types that express in higher intensities our markers :

```{r IC1, echo=FALSE}
if (!exists("cell_markers")){
  cell_markers <- loadRDS( "./data_BB/cell_markers_keren.rds")
}

ICMarkers <- cell_markers%>% dplyr::select(patient_id, cell_id, cell_type,Lag3,B7H3, IDO, PD1, PD.L1,OX40,Vimentin,Ki67,EGFR)
#%>% filter(cell_type =='CD4-T')
```

Now, let's take a look with a violin plot for each marker how high the cell types express It :

```{r IC2,echo = TRUE}
#ICM_m <- melt(ICMarkers[3:4])
ggplot(ICMarkers,aes(color = cell_type))+
    geom_violin(aes(x= cell_type, y = PD1),trim=TRUE,show.legend = FALSE)+
    theme(axis.text.x=element_text(angle=45,hjust=0.8,vjust=0.5))+
    coord_cartesian(ylim = c(-10,30))+
    labs(title = "    Boxplot of marker Lag3 intensity for each cell type ")+
    xlab ("") + ylab("intensity (Arbitrary Unit)")



```
  The violin plot doesn't clearly show us what are the cell types that express the most a marker such as Lag3. Let's try to plot points by cell types for each Immune checkpoint marker. 

```{r IC3,echo= FALSE}
#ic_markers <- list('Lag3','B7H3','IDO',PD1,PD.L1,OX40)
ic_markers <- ICMarkers[c('Lag3','B7H3','IDO','PD1','PD.L1','OX40')]#[4:9]
#par(mfrow(3,2))
plots <- list()
for(i in c('Lag3','B7H3','IDO','PD1','PD.L1','OX40','Vimentin','Ki67','EGFR')){
fig <- ggplot(ICMarkers,aes(color = cell_type))+
  geom_point(aes_string(x='cell_type', y =i),show.legend=FALSE)+
  theme(axis.text.x=element_text(angle=45,hjust=0.8,vjust=0.5))+
  coord_cartesian(ylim = c(0,40))+
  labs(title = paste("     Plot of marker ",i, " intensity for each cell type "))+
  xlab ("") + ylab("intensity (Arbitrary Unit)")
print(fig)
plots[[i]] <- fig  
}
#cowplot::plot_grid(plotlist = plots, nrow = 2)
#multiplot(plots,ncol = 3, nrow = 2)
```


```{r IC4, echo=TRUE}
CellsSitesIC <- left_join(cells_per_sites,ICMarkers,by = c('cell_id','patient_id'))%>%
  dplyr::select(-(cell_type.y))%>%
  dplyr::rename(cell_type=cell_type.x)
SummaryICSites <- CellsSitesIC %>% 
  group_by(patient_id,site_id,cell_type) %>%
  summarise(avLag3 = mean(Lag3),
            avOX40 = mean(OX40),
            avPD1 = mean(PD1),
            avPDL1 = mean(PD.L1),
            avIDO = mean(IDO),
            avB7H3 = mean(B7H3))

#SummaryICSites  <- SummaryICSites %>%mutate(avLag3= replace(avLag3,cell_type==,NA))

#### Lag3 marker
Lag3CD4T <- SummaryICSites%>% 
  group_by(patient_id,site_id)%>%
  filter(cell_type=='CD4-T')%>%
  dplyr::select(-c(cell_type,avOX40,avPD1,avPDL1,avIDO,avB7H3))%>%
  dplyr::rename(Lag3CD4 = avLag3)

Lag3CD8T <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='CD8-T')%>%
  dplyr::select(site_id,patient_id,avLag3)%>%
  dplyr::rename(Lag3CD8 = avLag3)

Lag3Tregs <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='Tregs')%>%
  dplyr::select(site_id,patient_id,avLag3)%>%
  dplyr::rename(Lag3tregs = avLag3)

#MARKERSic1 <- left_join(sites_PCA,Lag3Tregs,by=c('patient_id','site_id'))
###PD1 marker 
PD1CD4 <- SummaryICSites%>% group_by(patient_id,site_id)%>%
  filter(cell_type=='CD4-T')%>%
  dplyr::select(site_id,patient_id,avPD1)%>%
  dplyr::rename(PD1cd4 = avPD1)

PD1CD8 <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='CD8-T')%>%
  dplyr::select(site_id,patient_id,avPD1)%>%
  dplyr::rename(PD1cd8 = avPD1)


#### PD L1 marker
PDL1Ker <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='Keratin-positive tumor')%>%
  dplyr::select(site_id,patient_id,avPDL1)%>%
  dplyr::rename(PDL1ker = avPDL1)

#### OX40 marker
OX40CD8 <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='CD8-T')%>%
  dplyr::select(site_id,patient_id,avOX40)%>%
  dplyr::rename(OX40cd8 = avOX40)

OX40CD4 <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='CD4-T')%>%
  dplyr::select(site_id,patient_id,avOX40)%>%
  dplyr::rename(OX40cd4 = avOX40)

OX40Tregs <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='Tregs')%>%
  dplyr::select(site_id,patient_id,avOX40)%>%
  dplyr::rename(OX40tregs = avOX40)

OX40Ker <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='Keratin-positive tumor')%>%
  dplyr::select(site_id,patient_id,avOX40)%>%
  dplyr::rename(OX40ker = avOX40)

#### B7H3 marker
B7H3Ker <- SummaryICSites%>%
  group_by(patient_id,site_id)%>%
  filter(cell_type=='Keratin-positive tumor')%>%
  dplyr::select(site_id,patient_id,avB7H3)%>%
  dplyr::rename(B7H3ker = avB7H3)
#### IDO marker

```




```{r IC5, echo=TRUE}


## Sampling sites: gaussian counts (proportions)
sites_PCA <- read.csv("./data_BB/sites_pca_matrix_gaussian.csv",sep=',',header=TRUE)

## Sampling sites: absolute counts
#sites_PCA <-read.csv('/srv/mfs/hausserlab/anissa.el/building_blocks_old/sites_pca_matrix.csv',sep=',',header=TRUE)

sitesMarkersICperCellType <- left_join(sites_PCA,Lag3CD4T,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,Lag3Tregs,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,Lag3CD8T,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,PD1CD4,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,PD1CD8,by=c('patient_id','site_id'))

sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,PDL1Ker,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,OX40CD8,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,OX40CD4,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,OX40Tregs,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,OX40Ker,by=c('patient_id','site_id'))
sitesMarkersICperCellType <- left_join(sitesMarkersICperCellType,B7H3Ker,by=c('patient_id','site_id'))

sitesMarkersICperCellType <- sitesMarkersICperCellType%>%replace(., is.na(.), 0)

plot_ICMarkers <- function(marker_type){
  to_col <- sitesMarkersICperCellType$marker_type
  fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
                 y = sitesMarkersICperCellType$PC1,
                 z = sitesMarkersICperCellType$PC3,
                 type="scatter3d",
                 mode="markers",
                 mode = 'text',
                 color=~to_col)

  fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                     yaxis = list(title = 'PC2'),
                                     zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites')
  fig
}
### PLOT PCA SITES

#plot_ICMarkers(Lag3tregs)
```

```{r archetypesIC, echo=TRUE}
library(reticulate)

nbArch = 4
data = as.matrix(sites_PCA[,c("PC1","PC2","PC3")])

py_run_string("import numpy as np")
py_run_string("import sys")
py_run_string("sys.path.insert(1, '/srv/mfs/hausserlab/fabio/data_analysis/src/utils')")
py_run_string("from archetypes import ArchetypalAnalysis")
py_run_string("import pandas as pd")
py_run_string("def compute_archetypes(data,nbArch):
    pc3dWagner = data
    AA = ArchetypalAnalysis(n_archetypes = nbArch, 
                          tolerance = 0.001, 
                          max_iter = 200, 
                          random_state = 0, 
                          C = 0.0001, 
                          initialize = 'random',
                          redundancy_try = 30)
    AA.fit_transform(pc3dWagner)
    return AA.alfa")
ArchsAlfas <- t(py$compute_archetypes(data,as.integer(nbArch)))
sitesMarkersICperCellType$Arch1 <- ArchsAlfas[,1]  
sitesMarkersICperCellType$Arch2 <- ArchsAlfas[,2]
sitesMarkersICperCellType$Arch3 <- ArchsAlfas[,3]
sitesMarkersICperCellType$Arch4 <- ArchsAlfas[,4]
colnames(ArchsAlfas) <- c("Arch1","Arch2","Arch3","Arch4")
```


```{r plotsArchIC, echo=TRUE}
library(ggpubr)
sitesMarkerAA <- pivot_longer(sitesMarkersICperCellType%>%dplyr::select(-c(PC1,PC2,PC3)),#PD1cd4,site_id,Arch1,Arch2,Arch3,Arch4
                                      cols = c("Arch1","Arch2","Arch3","Arch4"),
                                      names_to = "archetype",
                                      values_to = "proportion")%>%pivot_longer(cols = c("Lag3CD4","Lag3tregs","Lag3CD8","PD1cd4","PD1cd8","PDL1ker","OX40cd8","OX40cd4","OX40tregs","OX40ker","B7H3ker"),
                                                                               names_to="marker",
                                                                               values_to="intensity")

ggplot(data=sitesMarkerAA,aes(x = proportion, y= intensity,color =marker ))+
  #scale_y_continuous(trans="log10")+
  facet_wrap(~archetype,ncol=2,labeller = "label_value")+
  geom_point()#+
  #stat_cor(method="spearman")
  #geom_smooth()  

# summary(lm(PD1cd4 ~ Arch1 + Arch3 + Arch3*Arch4+0,data=sitesMarkersICperCellType))
# summary(lm(PD1cd4 ~ Arch1*Arch2+0,data=sitesMarkersICperCellType))
# summary(aov(PD1cd4 ~ Arch3*Arch4+0,data=sitesMarkersICperCellType))
cellMarkers.sites <- right_join(cell_markers,cells_per_sites,by=c("cell_id","patient_id","cell_type"))

NB_PATIENTS <- 40
sites_PCA2 <-cbind(sites_PCA,ArchsAlfas)#nb patients = 40

cellMarkers.sites2 <- left_join(cellMarkers.sites,sites_PCA2, by = c("patient_id","site_id"))
cellMarkers.sites3 <- cellMarkers.sites2 %>%pivot_longer(cols = c("Background","dsDNA","Vimentin","SMA","B7H3","FoxP3","Lag3","CD4","CD16","CD56","OX40","PD1","CD31","PD.L1","EGFR","Ki67","CD209","CD11c","CD138","CD163","CD68","CSF.1R","CD8","CD3","IDO","Keratin17","CD63","CD45RO","CD20","p53","Beta.catenin","HLA.DR","CD11b","CD45","H3K9ac","Pan.Keratin","H3K27me3","phospho.S6","MPO", "Keratin6","HLA_Class_1"),
                                                        names_to = "marker",
                                                        values_to ="intensity")%>%pivot_longer(cols = c("Arch1", "Arch2", "Arch3", "Arch4"), names_to = "archetype", values_to ="proportion" )


ggplot(data = cellMarkers.sites3%>%filter(cell_type == "CD4-T" & marker %in% c("PD1","FoxP3","PD.L1","Lag3","IDO","B7H3")), aes(x = intensity, y = proportion,color = marker))+
  facet_wrap(~archetype,ncol=2, labeller="label_value")+
  geom_point()


```


```{r probsArchs, echo=TRUE}
#ggplot(data = sitesMarkerAA, aes(x = archetype,y = proportion))+
#  geom_col
# sites.ICMarkersArch<- sitesMarkersICperCellType%>%mutate(archetype = ifelse(Arch1>=0.8, "ARCH1",
#                                                       ifelse(Arch2>=0.8,"ARCH2",
#                                                              ifelse(Arch3>=0.8,"ARCH3",
#                                                                     ifelse(Arch4>=0.8,"ARCH4",NA)))))%>% filter(!is.na(archetype))
# anova (lm(PD1cd4 ~ archetype,data = sites.ICMarkersArch))
#anova(lm(PD1cd4~ Arch1 + Arch3 + Arch4 + Arch3*Arch4,data = sitesMarkersICperCellType))

#anova(lm(PD1cd4~ Arch2 + Arch3 + Arch4 + Arch3*Arch4,data = sitesMarkersICperCellType))
plot_IC_markers_archs <-function(cellType){
 plot1 <- ggplot(data = cellMarkers.sites3%>%filter(cell_type == cellType & marker %in% c("PD1","FoxP3","PD.L1","Lag3","IDO","B7H3")), aes(x = intensity, y = proportion,color = marker))+
  facet_wrap(~archetype,ncol=2, labeller="label_value")+
  ggtitle(ct)+
  geom_point()
 #plot1
 ggsave(plot1,paste0("./plots/plotICSitesArchs_",ct,".pdf"))
 #return(plot1)
}
#rm(ct)

# plot_IC_markers_archs("CD4-T")
# plots <- c()
# for(ct in CELLTYPES){
#   p1 <- plot_IC_markers_archs(ct)
#   #plots <-rbind(plots,p1) 
#   #print(p1)
# }


```
Look at the cells within the site 
For example a T cell
observe the intensity of a certain T cell marker
Environment: this is the proportions of archetypes
Plot : x is the Intensity of the marker observed in a specific cell type
      y is the proportion of as certain TMEN of a site in which the cell is located
      points are all the cells
      
WARNING: plot hist of markers ==> is it a good marker (bi modal distr), if it is bimodal then it is good quality. Otherwise, we would find anyways correlations between markers and the alphas, because of the neighbouring cells or because of a potenital overlapping between cells ==> CyTOF at singke cell level, more precise in space, less overlapping issues a priori.

```{r cellsMarkersSites ,echo=TRUE}

```

```{r plotIC1, echo = TRUE}

Lag3_Tregs <- sitesMarkersICperCellType$Lag3tregs
sitesMarkersICperCellType <- sitesMarkersICperCellType%>%
  replace_na(list(Lag3tregs=0))
fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type="scatter3d",
               mode="markers",color =~Lag3_Tregs,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of Lag3 in Tregs')
fig

```


We can find some sites where Lag3 is expressed in Tregs near Building block 3 (known to contain exclusively tumor cells). We can also find some along the edge between archetype 3 and 4. 

```{r plotIC2, echo = TRUE}
Lag3_CD4 <- sitesMarkersICperCellType$Lag3CD4
sitesMarkersICperCellType <- sitesMarkersICperCellType%>%
  replace_na(list(Lag3tregs=0))

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,type="scatter3d",
               mode="markers",
               color =~Lag3_CD4,
               opacity=0.5,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of Lag3 in CD4 T cells')
fig
```

Let's see with the PD1 marker :

```{r plotIC3, echo = TRUE}
PD1_CD4 <- sitesMarkersICperCellType$PD1cd4
#sitesMarkersICperCellType <- sitesMarkersICperCellType%>%replace_na(list(Lag3tregs=0))
fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type = "scatter3d",
               mode = "markers",
               color =~PD1_CD4,
               marker = list(colorbar = list(title = "Intensity of PD1 marker in CD4 T cells"),
                           colorscale = "Viridis"),
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of PD1 in CD4 T cells')
fig

```

We can have the same conclusions as previously. 

#### PD L1 MARKER

This marker is an immune checkpoint expressed by the tumor cells. It recognizes PD1 which is a molecule expressed by T cells. This complex allows tumor cell survival. T cells don't trigger the killing of those cells when the immune checkpoint is recognized. Here, we'll see the sites that contain tumor cells that express the marker PD L1 :



```{r plotIC4, fig.width =15,fig.height = 12,echo = FALSE}
sitesMarkersICperCellTypeCut = apply(sitesMarkersICperCellType, 2, function(x) {
  cutOffs = quantile(x, c(.05, .95))
  x[x<cutOffs[1]] = cutOffs[1]
  x[x>cutOffs[2]] = cutOffs[2]
 return(x)
}) %>% as.data.frame()
sitesMarkersICperCellType[1:5,1:7]
sitesMarkersICperCellTypeCut[1:5,1:7]
plot(sitesMarkersICperCellType[,8], sitesMarkersICperCellTypeCut[,8])

PDL1_KeratinPositiveTumor <- sitesMarkersICperCellTypeCut$PDL1ker

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type = "scatter3d",
               mode = "markers",
               color =~PDL1_KeratinPositiveTumor,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of PD1 L1 in Keratin positive tumor cells')
fig

```

#### OX40 MARKER

This marker is expressed mainly in CD4 T cells, Tregs and can be found in some tumor cells.

```{r plotIC5, echo = FALSE}
OX40_CD4 <- sitesMarkersICperCellType$OX40cd4

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type="scatter3d",
               mode="markers",
               color =~OX40_CD4,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of OX40 in CD4 T cells')
fig

```


In CD4 T cells, there are very few sites with this marker expressed in high intensity. However, those rare sites are located always on the edge between building block 3 and building block 4. 
Let's check for Tregs :

```{r plotIC6, echo = TRUE}
OX40_Tregs <- sitesMarkersICperCellType$OX40tregs

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type="scatter3d",
               mode="markers",
               color=~OX40_Tregs,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'), 
                                   yaxis = list(title = 'PC2'), 
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of OX40 in Tregs')
fig

```




```{r plotIC7, echo = TRUE}
OX40_KeratinPositiveTumor <- sitesMarkersICperCellType$OX40ker

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type="scatter3d",
               mode="markers",
               color=~OX40_KeratinPositiveTumor,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of OX40 in Keratin-Positive Tumor cells')
fig

```


#### B7H3 MARKER

```{r plotIC8, echo = TRUE}
B7H3_KeratinPositiveTumor <- sitesMarkersICperCellType$B7H3ker

fig <- plot_ly(x = sitesMarkersICperCellType$PC2,
               y = sitesMarkersICperCellType$PC1,
               z = sitesMarkersICperCellType$PC3,
               type="scatter3d",
               mode="markers",
               color=~B7H3_KeratinPositiveTumor,
               mode = 'text')

fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                                   yaxis = list(title = 'PC2'),
                                   zaxis = list(title = 'PC3') ),
                      title = '3D PCA of sites with average intensity of B7H3 in Keratin-Positive Tumor cells')
fig

```
 There are very few sites expressing in high intensity this marker, so this is not relevant for our analysis. BUt we know that this marker is expressed somewhere and not in sites classified as pure building blocks.
 
 


### **Take-home messages**

* **Building block 1** :  B cells (no sub-type found)  + probably T helper cells (CD45RO)

* **Building block 2** : Macrophages M1 (CD68+), CD8 and CD4 cells (no sub-types found)

* **Building block 3** :  More tumor cells, but maybe Macrophages (M0 or M1)

* **Building block 4** (disorganized tissue): Mix of Macrophages (no M2 macrophages) CD8  and CD3 T cells 

* We need to refine the identification of sub-type with another set of markers

* Some building blocks might look like vertices from tumor pop analysis (METABRIC)

* We have to use more accurate tools to map our findings with Fabio's building blocks

**NEW ANALYSIS**

* No checkpoint proteins expressed in any of the building block : Those cell types don't represent any of the 4 building block found after archetypal analysis

* No high Ki67 intensity for tumor cells : maybe for the same reasons aforementioned

* Tumor cells don't always express the same markers in the building blocks (Different cytokeratin expressed)

**Immune Checkpoint markers in tumor sites ?**

