---
title: "Linear Regression on TMENs"
author: "Anissa El Marrahi"
date: "11/10/2021"
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**

## Preliminary Settings

```{r libs, echo=TRUE}
rm(list=ls())
gc()
library(tidyverse)
library(ade4)
library(factoextra)
#library(readxl)
library(reshape2)
library(plotly)
#library(GGally)
library(ggpubr)
library(ggridges)

```


```{r settings, echo=TRUE}
dirName <- dirname(rstudioapi::getSourceEditorContext()$path )
setwd(dirName)#setwd(dir="/scratch/anissa.el/ImmuneStates/wagner_analysis")
source("./scripts/R/Wagner_Keren_functions.r")
MacroMicroData <- readRDS(file="./outputs/MacroMicroData.rds")
SitesBBKeren <- readRDS(file = "./data/archetypesSitesKeren.rds")
metadataWagner <- readRDS(file = "./data/metadataBC.rds")

```


```{r data1, echo=TRUE}
Y <- as_tibble(MacroMicroData$CellAbundanceBC) 
Ywagner <- Y%>%
  filter(group == "wagner_macro2")%>%
  dplyr::select(-(group))%>%
  column_to_rownames(var = "sample_id")
Ykeren <- Y%>%
  filter(group=="keren_macro")%>%
  dplyr::select(-(group))%>%
  column_to_rownames(var="sample_id")
TMENs <- Y%>%
  filter(group == "keren_micro")%>%
  dplyr::select(-(group))# %>%column_to_rownames(var = "sample_id")
B <- t(as.matrix(TMENs%>%column_to_rownames(var = "sample_id")))
```

## **Microscopic constraints on Macroscopic cell abundance**

```{r pairs, echo= TRUE}
#Yw.pairs <- cbind(Ywagner%>%gather(key = "x.key",value = "x.val"),Ywagner%>%gather(key = "y.key",value = "y.val"))
  #pivot_longer(cols= c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
  #                     names_to="x",values_to="val")
#Ybb.pairs <- cbind(B%>%pivot_longer(cols= c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
#                       names_to="x.key",values_to="x.val"),
#                   B%>%pivot_longer(cols= c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
#                       names_to="y.key",values_to="y.val"))
#colNb <- length(unique(Yw.pairs$y.key))
#Yw.pairsToPlot <- expand(Yw.pairs,x.key,y.key,x.val,y.val)%>%
#  filter((x.key =="B"& y.key =="CD4-T")|(x.key=="EpithelialCells"& y.key == "Mesenchymal-like")|(x.key == "Macrophages" & y.key == "EndothelialCells"))
ggplot(Ywagner, aes(x = `CD8-T`, y =`CD4-T`)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
#      facet_wrap(x.key ~ y.key, ncol = colNb, scales = 'free', labeller = label_both)#+
  geom_point(Y%>%filter(group == "keren_micro"),
             mapping=aes(x= `CD8-T`, y=`CD4-T`,colour=sample_id),size=6)+
  geom_polygon(data = Y%>%filter(group=="keren_micro"), fill="skyblue", alpha=0.3)+
  scale_color_manual("sample_id",values = c("BB1" ="#D463DF",#pink
                                "BB2" = "#FF0000", #redc()
                                "BB3" = "#1E90FF", #dodgerblue
                                "BB4" = "#000000"))+
  xlab("CD4 T")+ylab("CD8 T")
ggsave("./figs/CD8VSCD4_TMENs.pdf", height = 3, width = 4)

```

```{r plotPairs2, fig.width=10,fig.height=11,echo=TRUE}

cell_types <- c("CD4-T","CD8-T","B","EpithelialCells")
#Ywagner2 <- Ywagner%>%pivot_longer(cols=c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
#                              names_to="cell_type",values_to="proportion")%>%
#  filter(cell_type%in%cell_types)%>%mutate(cell_type2=cell_type)%>%mutate(prop2=proportion)

# Ywagner.cells <- cbind(Ywagner%>%pivot_longer(cols= c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
#                        names_to="x.key",values_to="x.val")%>%
#   filter(x.key%in%cell_types),
#                    Ywagner%>%pivot_longer(cols= c(EndothelialCells,EpithelialCells,`Mesenchymal-like`,NK,B,DC,`CD4-T`,`CD8-T`,Macrophages,OtherImmune),
#                        names_to="y.key",values_to="y.val")%>%
#   filter(y.key%in%cell_types))
# 
# colNb <- length(unique(Ywagner.cells$y.key))
# 
# Ywagner.cells <- expand(Ywagner.cells,x.key,y.key,x.val,y.val)
# ggplot(Ywagner.cells)+
#   geom_point(aes(x=x.val,y=y.val))+
#   facet_grid(x.key ~ y.key, scales = 'free', labeller = label_both)#, ncol = colNb


pairs(rbind(Ywagner[,cell_types],t(B)[,cell_types]),
      upper.panel = NULL,
      pch = rep(c(19,19),c(nrow(Ywagner), nrow(B))),
      col = c(rep("steelblue",nrow(Ywagner)),c("#D463DF","#FF0000","#1E90FF","#000000")),
      cex = rep(c(1.5,2.5),c(nrow(Ywagner), nrow(B))))


## Compute the error (unit: %) in cells % due to mass cytometry 
# 1/ (average total nb of live cells in one tumor sample)
NB_PATIENTS <- 144
LIVE_CELLS_PROP <- 0.85
TOTAL_NB_CELLS <- 26000000
err <- round((1/((LIVE_CELLS_PROP*TOTAL_NB_CELLS)/NB_PATIENTS))*100,digits=7)
Y.log10 <- log10(Ywagner+err)
B.log10 <- log10(t(B) + 100 / (90^2))
 
# ## Log10 functions of percentages:
# ct.x <- "CD4-T"
# ct.y <- "CD8-T"
# 
# values <- seq(B.log10[1,ct.x],B.log10[2,ct.x],0.1)
# 
# log10cells <- function(values){
#   
# }
  

pairs(rbind(Y.log10[,cell_types],B.log10[,cell_types]),
      upper.panel = NULL,
      pch = rep(c(19,19),c(nrow(Ywagner), nrow(B))),
      col = c(rep("steelblue",nrow(Ywagner)),c("#D463DF","#FF0000","#1E90FF","#000000")),
      cex = rep(c(1.5,2.5),c(nrow(Ywagner), nrow(B))))



  

```

## **Linear Regression**

**Parameters of the linear regression:**
* **Variable to explain: Y matrix, macroscopic cellular abundance of tumors**
* **Explanatory variable: B matrix, microscopic cellular abundance of 4 TMENs (TLS, Inflammatory nich, cancer and fibrotic niche)**

### **Linear regression with no interaction** 

```{r lm, echo=TRUE}
#Y.pourri <- Y%>%filter(group == "wagner_macro1")%>%dplyr::select(-(group))%>%column_to_rownames(var = "sample_id")
res.lm <- linear_regression(Y = as.matrix(Ywagner), B = B)
Omega <- t(res.lm$OmegaT)
RsqValues <- as_tibble(res.lm$R2_values,rownames=NA)%>%mutate(patients="BC patients")
#OmegaT <- solve(t(B) %*% B) %*% t(B) %*% t(Ywagner)

```

**Assessing the linear regression with R-square values:**

```{r r2, echo= TRUE}

violinPlot_R_square(RsqValues,savePlot=FALSE,nameToSave=NA)

```

**Plot the TMENs weights in 3D space: how are the tumors organized in this space ?**


```{r omegaPlot, echo=TRUE}

fig1 <- plot_ly(x = Omega[,"BB1"],
                y = Omega[,"BB2"],
                z = Omega[,"BB3"],
                #color=~Omega[,"BB4"],
                type = "scatter3d",
                name = "macroscopic tumors",
                mode = "markers",
                #inherit=FALSE,
                showlegend = TRUE,
                marker = list(size =5))#color = "#6495ED",
planeCol <- rgb(0, 0, 255, max = 255, alpha = 125, names = "blue20")

fig1 <- fig1%>%add_mesh(x = c(1,0,0),
                      y = c(0,1,0),
                      z = c(0,0,1),
                      name = "BB plane",
                      facecolor = planeCol,
                      opacity = 0.3,
                      inherit = FALSE,
                      showlegend = TRUE)

fig1 <- fig1%>%layout(scene = list(xaxis = list(title = "BB1"), 
                                  yaxis = list(title = "BB2"), 
                                  zaxis = list(title = "BB3") ),
                       title = "BC tumors from Wagner dataset fitted in the Building Blocks space")
fig1

```



#### Comparing with micro data of Keren dataset 

```{r kerenlm, echo=TRUE}
lmKeren <- linear_regression(Y = as.matrix(Ykeren), B = B)
Omega.keren <- t(lmKeren$OmegaT)
R2val.keren <- as_tibble(lmKeren$R2_values,rownames=NA)%>%
  mutate(patients="BC patients")

violinPlot_R_square(R2 = R2val.keren,savePlot = FALSE,nameToSave = NA)

```



```{r plotlm2}
fig2 <- fig1%>%add_trace(x = Omega.keren[,"BB1"],
                         y = Omega.keren[,"BB2"],
                         z = Omega.keren[,"BB3"],
                         inherit=FALSE,
                         type="scatter3d",
                         name="microscopic data",
                         showlegend=TRUE,
                         mode="markers",
                         marker=list(color="#B22222",size=5))
fig2
```



#### **Interaction between inflammatory block and TLS**


### **Linear regression with interaction BB1:BB2** 

```{r linCoeff12, echo=TRUE}
OmegaInflTLS <- Omega[,c("BB1","BB2")]
#Omega<-as_tibble(Omega,rownames=NA)
pcaBB1BB2 <-dudi.pca(OmegaInflTLS,center=FALSE,scale=FALSE,scannf=FALSE, nf=2)
fviz_eig(pcaBB1BB2)
fviz_pca_biplot(pcaBB1BB2,col.var = 'indianred2',col.ind = 'steelblue2', repel = TRUE)

plotbb12 <- plot(x = log(pcaBB1BB2$c1$CS1%*%t(Omega[,"BB1"])),
                 y=log(pcaBB1BB2$c1$CS1%*%t(Omega[,"BB2"])),
                 xlab="log(TLS)",
                 ylab= "log(Inflamatory block)",
                 main= "log-weights of TLS and inflammatory block across BC patients")
plotbb12
coeffBB1BB2 <- pcaBB1BB2$co$Comp1 #/(sum(pcaBB1BB2$co$Comp1))
#Create new TMEN: Immune TMEN(BB1BB2) is a combination of TLS and Inflammatory block
BB1BB2 <- (coeffBB1BB2[1]*B[,1] + coeffBB1BB2[2]*B[,2])/(coeffBB1BB2[1]+coeffBB1BB2[2])
# New matrix of TMENs cell abundance
B2<- cbind(B,BB1BB2)
B2 <- B2[,c("BB1BB2","BB3","BB4")]
#colnames(Omega)[5] <- "BB1BB2"
```



```{r plotBB1BB2, echo = TRUE}
#equation1 <- lm(log10(Omega[,"BB2"])~log10(Omega[,"BB1"]))
log10bb1bb2 <- log10(Omega[,c("BB1", "BB2")]) %>% as_tibble %>% drop_na()
firstEV <- eigen(cov(log10bb1bb2))$vectors[,1]
slope.log <- firstEV[2]/firstEV[1]
intcpt.log <- mean(log10bb1bb2 %>% pull(BB2)) - mean(log10bb1bb2 %>% pull(BB1))
cov.xx <- cov(log10bb1bb2)["BB1","BB1"]
cov.yy <- cov(log10bb1bb2)["BB2","BB2"]
cov.xy <- cov(log10bb1bb2)["BB1","BB2"]
G <- nrow(log10bb1bb2)

# f <-function(alpha){
#   ((1+alpha^2)^((G-3)/2))/((cov.yy+cov.xx*alpha^2 -2*alpha*cov.xy)^((G-1)/2))
# }
# 
# integral.f <- integrate(f,lower=0.5,upper=2,subdivisions = 10^5)
# alpha.val <- seq(0.5,2,10^-6)
# #integrate(f,lower=0,upper=0.)
# #format(round(1.1234, 2), nsmall = 2)
# low.CI <- signif((0.005)*integrate(f,lower=0.5,upper=2,subdivisions = 10^6)$value,digits=5)
# high.CI <- signif((0.995)*integrate(f,lower=0.5,upper=2,subdivisions = 10^6)$value,digits=5)
#  
# funct1 <- as_tibble(cbind(alpha.val,f(alpha.val)))
# alpha1 <- funct1[funct1$V2==max(funct1$V2),]$alpha.val
# 
# if(!exists("./outputs/alphaValuesCI99.rds")){
#   for(i in 1:length(alpha.val)){
#   #print(integrate(f,lower=0,upper=alpha.val[i])$value)
#   if(signif(integrate(f,lower = 0.5,upper = alpha.val[i])$value,5)==low.CI){
#     print("yes")
#     low.alpha <- alpha.val[i]
#     #print(low.alpha)
#     next
#     }
#   if(signif(integrate(f,lower = 0.5,upper = alpha.val[i])$value,5)==high.CI){
#     #print("yass queen!")
#     high.alpha <- alpha.val[i]
#     break
#   }
#   #saveRDS(list("alpha.low99"=low.alpha,"alpha.up99"=high.alpha),"./outputs/alphaValuesCI99.rds")
# }
# }else{
#   CI99.alpha <- readRDS("./outputs/alphaValuesCI99.rds")
#   alpha.low <- CI99.alpha$alpha.low99
#   alpha.high <- CI99.alpha$alpha.up99}
# 
# 
# # i<-1
# # diff1<-1
# # while(diff1!=0){
# #   diff1 <- abs(low.CI-signif(integrate(f,lower = 0.5,upper = alpha.val[i])$value,5))
# #   alpha.test1 <- alpha.val[i]
# #   i<- i+1
# # }
# 
# # plot(x = alpha.val,y = f(alpha.val),
# #      pch = 19,cex = 0.2,
# #      xlab = "alpha",ylab = "posterior probability (alpha)",
# #      main = "distribution of posterior probability of alpha")
# # abline(v=as.numeric(funct1[funct1$V2 ==max(f(alpha.val)),"alpha.val"]),col="red")
# # abline(v=as.numeric(funct1[funct1$alpha.val ==high.alpha,"alpha.val"]),col="dodgerblue")
# # abline(v=as.numeric(funct1[funct1$alpha.val ==low.alpha,"alpha.val"]),col="dodgerblue")
# #polygon(list(x=c(as.numeric(funct1[funct1$alpha.val ==low.alpha,"alpha.val"]),as.numeric(funct1[funct1$alpha.val ==high.alpha,"alpha.val"])),
# #        y=c(f(low.alpha),f(high.alpha))),
# #        col="dodgerblue")
# 
# #a.eq1 <- coef(equation1)[1]
# funct1<- funct1%>%dplyr::rename(y = V2)
# #b.eq1 <- coef(equation1)[2]
# ggplot(data=funct1,aes(x=alpha.val,y=y))+
#   geom_point(size=0.2)+
#   geom_line(aes(x=low.alpha,y=y))+
#   annotate("text",x=low.alpha-0.5,y=low.CI-10^4,label=str(low.alpha))+
#   geom_line(aes(x=high.alpha,y=y))+
#   annotate("text",x=high.alpha+0.5,y=low.CI-10^4,label=str(high.alpha))+
# #  geom_area(data=funct1[funct1$alpha.val>=low.alpha & funct1$alpha.val<=high.alpha,],fill="mediumseagreen")+
#   xlab("alpha")+ylab("posterior probability density")

#PC1.TMENs.log10 <- data.frame(PC1log10.x=as.vector(log10(pcaBB1BB2$c1$CS1%*%t(Omega[,"BB1"]))),
#                              PC1log10.y=as.vector(log10(pcaBB1BB2$c1$CS1%*%t(Omega[,"BB2"]))))

#library(ecoflux)
library(sfsmisc)
#scientific_10x(corTest.macro$p.value, digits = 1)

text.eq1 <- paste0("log10(y) = ",round(slope.log,3),"log10(x) + ",round(intcpt.log,3))
corTest.macro <- cor.test(x=log10(as.data.frame(Omega))%>%drop_na()%>%dplyr::select(BB1)%>%unlist(),
                                       y=log10(as.data.frame(Omega))%>%drop_na()%>%dplyr::select(BB2)%>%unlist(),method="pearson")
#bquote(italic(10)^2== .(r2))

r1 <- as.numeric(round(corTest.macro$estimate,2))
str1 <- c("R = "," p = "," \n")
value2 <- pretty10exp(corTest.macro$p.value,digits=3)[[1]]
text.macro <- as.expression(bquote(.(str1[1]) ~ .(r1) ~  .(str1[2]) ~ .(value2) * .(str1[3])))#formatC(corTest.macro$p.value, format = "e", digits = 2)
corTest.micro <- cor.test(x=log10(as.data.frame(Omega.keren))%>%drop_na()%>%dplyr::select(BB1)%>%unlist(),
                                     y=log10(as.data.frame(Omega.keren))%>%drop_na()%>%dplyr::select(BB2)%>%unlist(),method="pearson")

text.micro <- as.expression(bquote(.(str1[3]) ~ .(str1[1]) ~ .(round(corTest.micro$estimate,2)) ~
                     .(str1[2]) ~ .(round(corTest.micro$p.value,3))))
text2 <- c(text.macro,text.micro)
intercept_b <-mean(as.vector(log10(as.data.frame(Omega))%>%drop_na()%>%dplyr::select(BB2)%>%unlist()))-mean(as.vector(log10(as.data.frame(Omega))%>%drop_na()%>%dplyr::select(BB1)%>%unlist()))
# #text.eq1
#annotate("text",x=0.0039, y=0.5,label = text.eq1,color="darkorange1")+
#stat_smooth()+
ggplot(data = as_tibble(Omega,rownames = NA),aes(x = BB1 ,y = BB2,color = "Macroscopic data \n  (Wagner et al)"))+
  geom_point()+
  scale_x_continuous(trans = 'log10') + scale_y_continuous(trans = 'log10')+
  #stat_cor(method="pearson",label.x = log10(.060), label.y =log10(0.01) ,cor.coef.name = c("R"),color= "darkorange1")+
  annotate("text",x=.075, y=0.01,label = text.macro,color="darkorange1")+#annotate("text",x=0.0039, y=0.5,label = text.eq1,color="darkorange1")+
  geom_abline(slope=1, intercept=intercept_b,color="darkorange1",linetype="dashed")+
  geom_point(data = as_tibble(Omega.keren,rownames=NA),aes(x = BB1,y = BB2,color = "Microscopic data\n (Keren et al)"))+
  #stat_cor(data = as_tibble(Omega.keren,rownames=NA),aes(x = BB1,y = BB2),method="pearson",label.x = log10(.09), label.y = log10(.1),cor.coef.name = c("R"),color="cornflowerblue")+
  annotate(geom="text",x = .09, y = .005 ,label=text.micro,color="cornflowerblue")+
  theme(legend.title=element_blank())+
  scale_color_manual(values=c("Microscopic data\n (Keren et al)"="cornflowerblue","Macroscopic data \n  (Wagner et al)"= "darkorange1"))+
    xlab("TLS niche weight (log10)")+ylab("Inflammatory niche weight (log10)")+
  labs(color="Tumors") 
ggsave("./figs/TMEN1VS2_WK.pdf", height = 3, width = 5)


```


Now, let's plot the coefficients found in the linear regression (PCA) of BB2 over BB1 (alpha is the coefficient that determines the linearity between BB2 and BB1. If alpha ==1,  BB2/BB1):

```{r newBB1BB2, fig.width=10,fig.height=11,echo=TRUE}
coefB2 = 10^intcpt.log / (10^intcpt.log + 1)
coefB1 = 1 / (10^intcpt.log + 1)

## below is correct assuming s
#coefB2 <-1-(1/10^(intcpt.log)) #firstEV[1]/(firstEV[1]+firstEV[2])
#coefB1 <- 1/10^(intcpt.log)#firstEV[2]/(firstEV[1]+firstEV[2])
#B.log10 <- B.log10 %>%add_row(coefB1*BB1 + coefB2*BB2)
#B.log10 <-t(as.data.frame(t(B.log10))%>%mutate(BB1BB2 =coefB1*BB1 + coefB2*BB2))
B.log10 <- log10(t(as.data.frame(t(10^B.log10))%>%mutate(BB1BB2 =coefB1*BB1 + coefB2*BB2)) )
pairs(rbind(Y.log10[,cell_types],B.log10[c("BB1BB2","BB3","BB4"),cell_types]),
      upper.panel = NULL,
      pch = rep(c(19,19),c(nrow(Ywagner), nrow(B))),
      col = c(rep("steelblue",nrow(Ywagner)),c("#6C18A7","#1E90FF","#000000")),#c("#D463DF","#FF0000","#1E90FF","#000000")
      cex = rep(c(1.5,2.5),c(nrow(Ywagner), nrow(B))))


```


```{r plotBoundaries,fig.width=12,fig.height=10,echo=TRUE}
# C1<- "B" #"CD4-T"
# C2 <- "EpithelialCells" # "CD8-T"
# TMEN1 <- "BB3"
# TMEN2 <- "BB1BB2"
colsTMENs <- list("BB1"="#D463DF",
                  "BB2"="#FF0000",#"BB1BB2" = "#6C18A7",
                  "BB3" = "#1E90FF",
                  "BB4" = "#000000")

library(MASS)

# x2 <- seq(B.log10[TMEN1,C1],B.log10[TMEN2,C1],.1)
# f2 <- function(x){
#   log10(((10^x -B["B","BB1"])/(B["B","BB3"]-B["B","BB1"]))*(B["EpithelialCells","BB3"]-B["EpithelialCells","BB1"])+B["EpithelialCells","BB1"])}

Kerr = 100 / (90^2) # Error from Keren dataset
B3 <- as.data.frame(B+Kerr)#%>%mutate(BB1BB2 =((coefB1*BB1)+(BB2*coefB2)))%>%dplyr::select(-c(BB1,BB2))


plot_tmens_boundaries <- function(B=B3,OmegaMatrix = Omega,Blog10=B.log10,Ylog10=Y.log10,cellsLabels = c("CD4-T","B"),colors.TMENs = colsTMENs){
  par(mfrow = c(1, 1))#par(mfrow = c(2, 3)) # grid of plots for 6 plots 
  par(oma = c(4, 4, 0, 0)) # make room (i.e. the 4's) for the overall x and y axis titles
  #par(mar = c(2, 2, 1, 1)) # make the plots be closer together
  
  # Create pairs of cell labels: our xy axes
  # Create pairs of TMENs: plot a line between them (with the function f)
  combs.cells <- combn(cellsLabels,2)
  comb.tmens <- combn(names(colors.TMENs),2)
  maxWeightsTMENS <- right_join(as.data.frame(OmegaMatrix)%>%rownames_to_column(var="sample_id"),
                                as.data.frame(Ylog10)%>%drop_na()%>%rownames_to_column(var="sample_id"),
                                by="sample_id")%>%
    filter(BB1 == max(BB1) | BB2 ==max(BB2) | BB3 ==max(BB3) | BB4 == max(BB4))%>%group_by(sample_id)%>%
    mutate(MaxWeight = ifelse(max(c(BB1,BB2,BB3,BB4))==BB1,"BB1",
                              ifelse(max(c(BB1,BB2,BB3,BB4))==BB2,"BB2",
                                     ifelse(max(c(BB1,BB2,BB3,BB4))==BB3,"BB3","BB4"))))%>%
    column_to_rownames(var="MaxWeight")
  #print(f2(x.vals,coordX,coordY,b1,b2))
  for(j in 1:ncol(combs.cells)){
    coordX <- combs.cells[1,j] # Define the x-y axes as combinations of cell labels
    coordY <- combs.cells[2,j]
    
    
    #Function:  TMEN1 = f(TMEN2) in log10 scale
    f2 <- function(x,coordX,coordY,b1,b2){
      log10(((10^x -B[coordX,b2])/(B[coordX,b1]-B[coordX,b2]))*(B[coordY,b1]-B[coordY,b2])+B[coordY,b2])}
    
    #FUnction for max weights TMENmax1 = g(TMENmax1)
    g2 <- function(x,coordX,coordY,b1,b2){
      log10(((10^x -maxWeightsTMENS[b2,coordX])/(maxWeightsTMENS[b1,coordX]-maxWeightsTMENS[b2,coordX]))*(maxWeightsTMENS[b1,coordY]-maxWeightsTMENS[b2,coordY])+maxWeightsTMENS[b2,coordY])}
    
    #X boundaries
    #Xmin <- round(min(as.data.frame(rbind(Ylog10,Blog10)[,coordX])%>%drop_na()),3)
    #Xmax <- round(max(as.data.frame(rbind(Ylog10,Blog10)[,coordX])%>%drop_na()),3)
    Xmin <- min(as.data.frame(rbind(Ylog10,Blog10)[,coordX])%>%drop_na())
    Xmax <- max(as.data.frame(rbind(Ylog10,Blog10)[,coordX])%>%drop_na())
    
    #Y boundaries 
    #Ymin <- round(min(as.data.frame(rbind(Ylog10,Blog10)[,coordY])%>%drop_na()),3)
    #Ymax <- round(max(as.data.frame(rbind(Ylog10,Blog10)[,coordY])%>%drop_na()),3)
    Ymin <- min(as.data.frame(rbind(Ylog10,Blog10)[,coordY])%>%drop_na())
    Ymax <- max(as.data.frame(rbind(Ylog10,Blog10)[,coordY])%>%drop_na())
    newCoordX <- set_cells_names(cell_types[cell_types==coordX])
    newCoordY <- set_cells_names(cell_types[cell_types==coordY])
    #Plot macroscopic tumor samples
    plot(x = Ylog10[,coordX],y = Ylog10[,coordY],pch = 19,col = "steelblue",
         xlim = c(Xmin,Xmax),
         ylim = c(Ymin,Ymax),
         xlab = paste("% of ",newCoordX,"(log10)"),ylab = paste("% of ",newCoordY,"(log10)"))
    #xVect <- unlist(as.data.frame(Ylog10%>%drop_na())[,coordX])
    #yVect <- unlist(as.data.frame(Ylog10%>%drop_na())[,coordY])
    #biv_kde <- MASS::kde2d(xVect, yVect)
    #contour(biv_kde,levels=c(0.05, 0.5), add = T)
    #axis(1,labels=coordX)
    #axis(2,labels=coordY)
    #Plot TMENs
    points(x = Blog10[names(colors.TMENs),coordX],y = Blog10[names(colors.TMENs),coordY],
           col = unlist(colors.TMENs),pch = 19,cex = 2)
    #Plot max weigths of TMENs in linear regression
    #points(x= maxWeightsTMENS[,coordX],y = maxWeightsTMENS[,coordY],col="red",pch=9,cex=2)
    
    #Plot line between two TMENs
    for(k in 1:ncol(comb.tmens)){
      TMEN1 <- comb.tmens[1,k]
      TMEN2 <- comb.tmens [2,k]
      ## If both points have th same x value, then draw a straight line between them:
      #if(round(Blog10[TMEN1,coordX],3)==round(Blog10[TMEN2,coordX],3)){
      if ( abs(Blog10[TMEN1,coordX] - Blog10[TMEN2,coordX]) < .05 ) {
        #print("OK")
        points(x = c(Blog10[TMEN1,coordX],Blog10[TMEN1,coordX]),
               y = c(Blog10[TMEN1,coordY],Blog10[TMEN2,coordY]),
           type = "l",lty = 1)
      }
      else{
        if(Blog10[TMEN1,coordX]>Blog10[TMEN2,coordX] ){ #| maxWeightsTMENS[TMEN1,coordX]>maxWeightsTMENS[TMEN2,coordX]){#
          #print("here")
          old.TMEN1 <- TMEN1
          TMEN1 <- TMEN2 
          TMEN2 <- old.TMEN1
        }
        x.vals <- seq(from=Blog10[TMEN1,coordX],
                      to=Blog10[TMEN2,coordX],
                      len=100)
        # x.vals2 <- seq(from=maxWeightsTMENS[TMEN1,coordX],
        #               to=maxWeightsTMENS[TMEN2,coordX],
        #               len=100)
        # if(is.na(f2(x.vals,coordX,coordY,b1=TMEN1,b2=TMEN2))){
        #   print("STOP!")
        #   print(TMEN1)
        #   print(TMEN2)
        #   print(coordX)
        #   print(coordY)}
        points(x = x.vals,y = f2(x.vals,coordX,coordY,b1=TMEN1,b2=TMEN2),
             type = "l",lty = 1)
        # points(x = x.vals2,y = g2(x.vals2,coordX,coordY,b1=TMEN1,b2=TMEN2),
        #      type = "l",lty = 2,col="red")
      }
    }
  }
 

} 


pdf("./figs/plotTMENsBound_CD4vsB.pdf", height=6, width=7)
plot_tmens_boundaries(B = B3,cellsLabels=c("CD4-T","B"),colors.TMENs = colsTMENs)#plot_tmens_boundaries(coordX=C1,coordY=C2,b1=TMEN1,b2=TMEN2)
dev.off()
#plot_tmens_boundaries(coordX=C1,coordY=C2,b1="BB4",b2="BB1")
#plot_tmens_boundaries(coordX=C1,coordY=C2,b1="BB3",b2="BB2")

```



**Plot of TMENs weights with "interaction" between TLS and Inflammatory TMEN**

```{r lm_bb1bb2,echo=TRUE}
lm.inter <- linear_regression(Y = Ywagner,B=as.matrix(as.data.frame(B+Kerr)%>%mutate(BB1BB2 =((coefB1*BB1)+(BB2*coefB2)))%>%dplyr::select(-c(BB1,BB2))))
Omega.inter <-t(lm.inter$OmegaT) 
R2val.inter <- as_tibble(lm.inter$R2,rownames="patient_id")%>%mutate(patients="128 patients")%>%column_to_rownames(var="patient_id")
#Plot distribution of R square values
violinPlot_R_square(R2 = R2val.inter,savePlot=TRUE,nameToSave="VlnPlotR2_lm")
```



```{r fig3,echo=TRUE}

metadataBC.wagner <- metadataWagner%>%filter(`Health Status` =="Breast cancer patient")%>%
  mutate(Patient =paste0("T_",Patient))

left_join(Omega.inter%>%as_tibble(rownames=NA)%>%rownames_to_column(var="Patient"),metadataBC.wagner,by="Patient")
## Plot Omega values with coupling with TLS and Inflammatory block 
fig3 <- plot_ly(x = Omega.inter[,"BB1BB2"],
                y = Omega.inter[,"BB3"],
                z = Omega.inter[,"BB4"],
                #color=~Omega[,"BB4"],
                type = "scatter3d",
                name = "macroscopic tumors",
                mode = "markers",
                #inherit=FALSE,
                showlegend = TRUE,
                marker = list(size = 5))#color = "#6495ED",
planeCol <- rgb(0, 0, 255, max = 255, alpha = 125, names = "blue20")

fig3 <- fig3%>%add_mesh(x = c(1,0,0), #c(max(Omega.inter[,"BB1BB2"]),0,0),
                        y = c(0,1,0), #c(0,max(Omega.inter[,"BB3"]),0),
                        z = c(0,0,1), #c(0,0,max(Omega.inter[,"BB4"])), 
                        name = "BB plane",
                        facecolor = planeCol,
                        opacity = 0.3,
                        inherit = FALSE,
                        showlegend = TRUE)

fig3 <- fig3%>%layout(scene = list(xaxis = list(title = "TLS-Inflammatory TMEN"), 
                                   yaxis = list(title = "cancer"), 
                                   zaxis = list(title = "fibrotic")),
                       title = "BC tumors from Wagner dataset fitted in TMENs space")
#pdf(file = "./figs/testPDF.pdf", dpi)
fig3
#dev.off()

```

### **Assessment of linear regression: residuals and predicted values**


```{r obsVSpred,echo=TRUE}

plot_obs_VS_pred_patient <- function(Y, B, R2, Omega,error = 0,log10Scale = FALSE,savePlot = FALSE,namePlot = " "){
  Y <- Y + error
  B <- B + error
  R2 <- R2 %>%rownames_to_column(var = "names")

  colnames(Y) <- colnames(Y)%>%set_cells_names()#change cell labels for plot
  rownames(B) <- rownames(B)%>%set_cells_names()#change cell labels for plot
  #print(Y)
  plots <- list()
  #names(plots) <- rownames(Y)
  names_patients <- rownames(Y)
  #for (i in rownames(Y)){
  plots <- lapply(names_patients, function(x){
    avPatient <- R2[x, ] #R2[which(R2$names == x),value]
    namePlot=x
    #Create df of observed cell% (Y) and predicted cell % (B%*%t(Omega))
    ObsVSPred <- data.frame(Observed = as.vector(t(Y[x,])),#t(Y)[,avPatient$names],
                             Predicted=B%*%t(Omega)[,x])#t(B%*%t(Omega)[,avPatient$names]))
    print(ObsVSPred)
    #ObsVSPred <- data.frame(Observed = t(Y),
    #                        Predicted=(B%*%t(Omega)))
    ObsVSPred<- as_tibble(ObsVSPred,rownames=NA)%>%
      rownames_to_column(var="cell_type")%>%
      mutate(cell_type = set_cells_names(cell_type))# set same cell labels

    #Scatterplot of cell % (Observed VS predicted)
    
    r2 <- round(as.numeric(avPatient$value),2)
    label1 <- as.expression(bquote(italic(r)^2== .(r2)))
    #print(label1)
    p <- ggplot(data = ObsVSPred,aes(x = Predicted,y = Observed))+
      geom_point()+
      ggrepel::geom_text_repel(aes(label = cell_type),max.overlaps=13)+
      geom_abline(slope = 1, intercept = 0) + 
      xlab("% predicted tumor cellular composition")+
      ylab("% observed tumor cellular composition ")+
    annotate("text",x =5,y=40,label=avPatient$names)+
    annotate("text", x = 5, y = 33, label = label1)#expression(paste(italic(R)^2, " = ", format(r2, digits = 2)))
    #plot
    
    ## Saving fig in pdf file in figs directory
    if(savePlot==TRUE){
      figName  <- paste0("./figs/",namePlot,".pdf") #add the name of the measure (mean or median) in the name of the file
       ggsave(figName,p, height = 3, width = 4)
     }     
  })
 names(plots) <- names_patients
  return(plots)
}

plots <-plot_obs_VS_pred_patient(Y = Ywagner, B = B2,Omega = Omega.inter, 
                          R2 = R2val.inter,error = 0, savePlot=TRUE,log10Scale = F)
#plots
# for (i in names(plots)){
#   ggsave(paste0("./figs/ObsVSPreds/ObsVSPred_",i,".pdf"),plots[[i]],height=4,width=5)
# }

#lapply(names(plots),function(x) {ggsave(paste0("./figs/ObsVSPreds/ObsVSPred_",x,".pdf"),plots[[x]],height=4,width=5)})

#map2(plots, names(plots),function(x, name){ggsave(paste0("./figs/ObsVSPreds/ObsVSPred_",name,".pdf"),x,height=4,width=5)})

```

#### **Residuals analysis**

We will analyze the nature of residuals. Residuals are just (observed values - predicted value). Here, as we performed a linear regression on macroscopic cellular composition, the observed value would be the cell percentages across tumors. We are going to assess the assumptions of our linear model. Residuals are considered as constant across fitted values and follow a gaussian distribution (centered to 0).

```{r res1, echo=TRUE}
#residuals.cd4 <- as.data.frame(res.lm$residual)$`CD4-T`
CELL.LABELS <- c("CD4-T","CD8-T","B","EpithelialCells")
for(cellLabel in CELL.LABELS){
  y.fit <- (Omega%*%t(B))[,cellLabel]
  residuals.cellType <- (Ywagner[,cellLabel]-y.fit)
  ## Testing normality of residuals
  pval <- shapiro.test(residuals.cellType)$p.value
  #print(var(residuals.cellType))
  ## Checking hetero/homoscedasticity of residuals
  plot(x = y.fit,y = residuals.cellType,
       xlab = paste("fitted %",cellLabel),
       ylab = paste("residuals for %",cellLabel),
       pch = 19,col = "steelblue")
  grid(nx = 8, ny = 6)
  abline(a = 0,b = 0,col = "black",lty = 10)
  text(x = 6,y = 14,labels = paste("Shapiro test (residuals)\n p value:",formatC(pval, format = "e", digits = 2)))
} 


```


#### **Observed versus predicted values**

```{r ObsVSPred2,echo=TRUE}
## Create a dataframe of predicted versus observed values
CELL.LABELS <- c("CD4-T","CD8-T","B","EndothelialCells","EpithelialCells","Mesenchymal-like","DC","Macrophages","OtherImmune","NK")
ObsVSPred.cells <- inner_join(as_tibble(Ywagner[,all_of(CELL.LABELS)]+err,rownames=NA)%>%rownames_to_column(var="sample_id")%>%
                                pivot_longer(cols=all_of(CELL.LABELS),names_to="cell_type",values_to="observed_prop"),
                        as_tibble((Omega%*%t(B))[,all_of(CELL.LABELS)]+err,rownames=NA)%>%rownames_to_column(var="sample_id")%>%
                          pivot_longer(cols=all_of(CELL.LABELS),names_to="cell_type",values_to="predicted_prop"),
                        by=c("sample_id","cell_type"))%>%mutate(cell_type=set_cells_names(cell_type))

ggplot(data=ObsVSPred.cells)+
  geom_point(aes(x = observed_prop,y = predicted_prop))+
  scale_x_continuous(trans='log10') + scale_y_continuous(trans="log10")+
  facet_wrap(~cell_type,ncol=5, labeller = "label_value")+
  geom_abline(slope = 1, intercept = 0,linetype = "dashed")+
  xlab("observed %") + ylab("predicted %")+
  theme(axis.text.x = element_text(angle = 45, vjust = .2))
ggsave("./figs/ObsVSPredictedCells.pdf", height = 4, width = 8)
# ggplot(data=ObsVSPred.cells)+
#   geom_point(aes(x=observed_prop ,y=predicted_prop))+
#   facet_wrap(~cell_type,ncol=5, labeller = label_both)+
#   geom_abline(slope = 1, intercept = 0,linetype = "dashed")+
#   theme(axis.text.x = element_text(angle = 45, vjust = .2))


```



### **Distribution of TMENs across patients**

```{r densities, echo=TRUE}
#theme_set(theme_ridges())
#ggplot(BBscores, aes(y = proportion)) +
#  geom_density_ridges(aes(x = building_block,y = proportion, fill = proportion))# +
#  scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07","#FC4E08"))
##Creating same dataset as BBscores but here with the building block BB1:BB2 (linear comb of BB1 and BB2):
BBscores.inter <- as_tibble(Omega.inter,rownames = NA)%>%
  rownames_to_column(var="sample")%>%
  pivot_longer(c(BB1BB2,BB3,BB4),names_to = "building_block",values_to = "props")

gg1 <-ggplot(BBscores.inter, aes(y =  building_block)) +
geom_density_ridges(
    aes(x = props, fill = building_block), 
    alpha = .8, color = "white", from = 0, to = 1)+
  labs(x = "proportion",
       y ="building block",
       title = "density of building blocks proportions across BC samples (Wagner)")

gg1

```

